State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources

### 2020-09-23 analysis/paper/paper.Rmd
added code to refine the figures, and save black and white and color versions of them as .tiff files, as required by NAFMJ. Added all .tiff files to Git LFS.

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      # fig_height: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Accurate estimates of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin. Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate. We tested the approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time. Our results demonstrate that the model produced unbiased estimates across all tested scenarios. We also applied this model to data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring-summer run Chinook Salmon and steelhead from spawn years 2010-2019. 
keywords: |
  escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 600
)
```

@@

Funding for this study and development of the STADEM model was partially provided by the Bonneville Power Administration under project 2003-017-00. Special thanks to Darren Ogden and staff at the Lower Granite Dam adult trap for their hard work and diligent data collection. Thank you to personnel at the Idaho Department of Fish and Game, particularly Paul Bunn, Tim Copeland, and Bill Schrader, for providing access to data from the adult fish trap and conceptualizing methods, and to Matthew Campbell and staff at the Eagle Fish Genetics Laboratory for analyzing genetic samples. Special thanks to Columbia Basin Research staff and the Columbia River Data Access in Real Time (DART) application, and Susannah Iltis in particular. And finally, thank you to Rick Orme for his contributions to the development of STADEM, and all the other folks who have contributed through productive critique and conversations.

@@

@@

```{r stadem-examp-fig, fig.width=3.5, fig.asp = 2, fig.cap = "Schematic showing the STADEM model framework. (A) shows the posterior of the estimate of fish crossing the dam while the window is open (dashed line shows observed window counts). That estimate is divided by the nighttime passage rate (B). The total fish is then discounted by the re-ascension rate to estimate unique fish (C). Those unique fish are then multiplied by the proportions by origin (D), to estimate unique fish by origin."}

@@

# choose appropriate color scale for color vs. black & white
p1 = ggplot(data.frame(x = c(-5, 8)),
            aes(x)) +
  stat_function(fun = dnorm,
                args = list(mean = 4, sd = 0.4),
                aes(color = 'Daytime Fish'),
                lwd = 2) +
  stat_function(fun = dnorm,
                args = list(mean = 6, sd = 0.55),
                aes(color = 'Total Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = 5, sd = 0.6),
                aes(color = 'Unique Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique Wild Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique Hatchery Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique HNC Fish'),
                geom = 'blank') +
  scale_color_brewer(palette = "Spectral") +
  # scale_color_grey(start = 0.9,
  #                  end = 0.3) +
  geom_vline(xintercept = 4,
             linetype = 2) +
  theme_pubr(legend = "bottom") +
  guides(color = guide_legend(nrow = 3,
                              title.position = 'top')) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y = element_blank()) +
  labs(x = 'Escapement',
       color = 'Fish Type',
       title = '   Daytime Fish') +
  coord_cartesian(xlim = my_xlim,
                  ylim = my_ylim)
# coord_fixed(ratio = 4,
  #             xlim = my_xlim,
  #             ylim = my_ylim)

@@

examp_p = ggarrange(p1, p2a, p3a, p4,
                    labels = 'AUTO',
                    ncol = 1,
                    nrow = 4,
                    align = 'v',
                    common.legend = T,
                    legend = 'bottom')

examp_p

@@

rel_bias_p
```

@@

ts_col
# ts_bw

@@

```{r night-reasc-diff-fig, fig.height=5, fig.cap = "Nighttime passage rate plotted against re-ascension rate on the logit scale, calculated from observed PIT tags for each week of spawn years 2010-2019. Colors correspond to different spawn years, while the size of each point is proportional to the window count that week. The dashed line is the 1-1 line."}

@@

# color version
rate_p_col = rate_p_base +
  scale_color_brewer(palette = "Paired") +
  geom_point(aes(size = win_cnt,
                 color = as.factor(Year))) +
  labs(color = "Spawn Year",
       size = "Window Count") +
  theme_pubr(legend = "left")

# black and white version
rate_p_bw = rate_p_base +
  geom_point(aes(size = win_cnt),
             shape = 21,
             fill = 'gray40',
             color = 'black',
             alpha = 0.3)

rate_p_col
# rate_p_bw
```

>>>>

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      # fig_height: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Accurate estimates of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin. Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate. We tested the approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time. Our results demonstrate that the model produced unbiased estimates across all tested scenarios. We also applied this model to data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring-summer run Chinook Salmon and steelhead from spawn years 2010-2019. 
keywords: |
  escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)

# set default values for saving alternative figures (color versions)
my_dpi = knitr::opts_chunk$get()$dpi
my_ext = "tiff"
```

@@

Funding for this study and development of the STADEM model was partially provided by the Bonneville Power Administration under project 2003-017-00. Special thanks to Darren Ogden and staff at the Lower Granite Dam adult trap for their hard work and diligent data collection. Thank you to personnel at the Idaho Department of Fish and Game, particularly Paul Bunn, Tim Copeland, and Bill Schrader, for providing access to data from the adult fish trap and conceptualizing methods, and to Matthew Campbell and staff at the Eagle Fish Genetics Laboratory for analyzing genetic samples. Special thanks to Columbia Basin Research staff and the Columbia River Data Access in Real Time (DART) application, and Susannah Iltis in particular. And finally, thank you to Rick Orme for his contributions to the development of STADEM, and all the other people who have contributed through productive critique and conversations.

@@

\listoffigures

@@

```{r stadem-examp-fig, fig.height = 5, fig.cap = "Schematic showing the STADEM model framework. (A) shows the posterior of the estimate of fish crossing the dam while the window is open (dashed line shows observed window counts). That estimate is divided by the nighttime passage rate (B). The total fish is then discounted by the re-ascension rate to estimate unique fish (C). Those unique fish are then multiplied by the proportions by origin (D), to estimate unique fish by origin."}

@@

# choose appropriate color scale for color vs. black & white
p1 = ggplot(data.frame(x = c(-5, 8)),
            aes(x)) +
  stat_function(fun = dnorm,
                args = list(mean = 4, sd = 0.4),
                aes(color = 'Daytime Fish'),
                lwd = 2) +
  stat_function(fun = dnorm,
                args = list(mean = 6, sd = 0.55),
                aes(color = 'Total Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = 5, sd = 0.6),
                aes(color = 'Unique Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique Wild Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique Hatchery Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique HNC Fish'),
                geom = 'blank') +
  scale_color_brewer(palette = "Spectral") +
  geom_vline(xintercept = 4,
             linetype = 2) +
  theme_pubr(legend = "bottom") +
  guides(color = guide_legend(nrow = 3,
                              title.position = 'top')) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y = element_blank()) +
  labs(x = 'Escapement',
       color = 'Fish Type',
       title = '   Daytime Fish') +
  coord_cartesian(xlim = my_xlim,
                  ylim = my_ylim)
# coord_fixed(ratio = 4,
  #             xlim = my_xlim,
  #             ylim = my_ylim)

@@

examp_col = ggarrange(p1, p2a, p3a, p4,
                    labels = 'AUTO',
                    ncol = 2,
                    nrow = 2,
                    align = 'v',
                    common.legend = T,
                    legend = 'bottom')

bw_scale = c(0.9, 0.1)
examp_bw = ggarrange(p1 +
                       scale_color_grey(start = bw_scale[1],
                                        end = bw_scale[2]), 
                   p2a +
                       scale_color_grey(start = bw_scale[1],
                                        end = bw_scale[2]), 
                   p3a +
                       scale_color_grey(start = bw_scale[1],
                                        end = bw_scale[2]), 
                   p4 +
                       scale_color_grey(start = bw_scale[1],
                                        end = bw_scale[2]),
                    labels = 'AUTO',
                    ncol = 2,
                    nrow = 2,
                    align = 'v',
                    common.legend = T,
                    legend = 'bottom')

examp_bw

# save black and white version
ggsave(paste0("../figures/stadem-examp-fig-bw.", my_ext),
       examp_bw,
       height = 5,
       width = 7,
       dpi = my_dpi)

# save color version
ggsave(paste0("../figures/stadem-examp-fig-col.", my_ext),
       examp_col,
       height = 5,
       width = 7,
       dpi = my_dpi)

@@

rel_bias_p

# save black and white version
ggsave(paste0("../figures/rel-bias-lgd-bw.", my_ext),
       rel_bias_p,
       height = 5,
       width = 3.5,
       dpi = my_dpi)

@@

# ts_col
ts_bw

# save black and white version
ggsave(paste0("../figures/jags-fit-fig-bw.", my_ext),
       ts_bw,
       height = 5,
       width = 7,
       dpi = my_dpi)

# save color version
ggsave(paste0("../figures/jags-fit-fig-col.", my_ext),
       ts_col,
       height = 5,
       width = 7,
       dpi = my_dpi)
```

@@

```{r night-reasc-diff-fig, fig.height=5, fig.cap = "Nighttime passage rate plotted against re-ascension rate on the logit scale, calculated from observed PIT tags for each week of spawn years 2010-2019. The size of each point is proportional to the window count that week. The dashed line is the 1-1 line."}

@@

# color version
rate_p_col = rate_p_base +
  scale_color_brewer(palette = "Paired") +
  geom_point(aes(size = win_cnt,
                 color = as.factor(Year))) +
  labs(color = "Spawn Year",
       size = "Window Count") +
  theme_pubr(legend = "left") +
  theme(axis.text = element_text(size = 10))

# black and white version
rate_p_bw = rate_p_base +
  geom_point(aes(size = win_cnt),
             shape = 21,
             fill = 'gray40',
             color = 'black',
             alpha = 0.3) +
  labs(size = "Window Count") +
  theme_pubr(legend = "bottom") +
  theme(axis.text = element_text(size = 10)) +
  guides(size = guide_legend(nrow = 2))

# rate_p_col
rate_p_bw

# save black and white version
ggsave(paste0("../figures/night-reasc-diff-fig-bw.", my_ext),
       rate_p_bw,
       width = 3.5,
       height = 7,
       dpi = my_dpi)

# save color version
ggsave(paste0("../figures/night-reasc-diff-fig-col.", my_ext),
       rate_p_col,
       width = 5,
       height = 7,
       dpi = my_dpi)
```
